---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Reproducible Research: Course Project 1
========================================================

**Read the file**

```{r}
# Download csv file from internet
rm(list = ls())
getwd()
setwd("C:/Users/Deepak/Documents")
directory <- "John Hopkins Data science/Reproducible Research/Week 2"
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
path <- paste(getwd(),directory,sep="/")
print(path)
filename <- "activity.zip"
fullname <- paste(path,filename,sep="/")
fullname
#download.file(url,destfile = fullname)
#dateDownloaded <- date()

## Unzip the files and save the list of files into data
actData <- unzip(fullname,list = TRUE)$Name
## actData

## Remove the variables not required from the workspace to clear the space
#rm(url)
#rm(path)
#rm(filename)
#rm(fullname)

## Rad the actData into R
actData <- read.csv(actData) #17568 obs. of 3 variables
str(actData)
head(actData)

```


**Process/Transform the data (if necessary) into a format suitable for your analysis**
```{r}

library(plyr)

actData$date<-as.Date(actData$date)

actSum<-ddply(actData, "date", summarize, totalSteps=sum(steps,na.rm = TRUE))
head(actSum)
str(actSum)
```


**Make a histogram of the total number of steps taken each day and save the plot as a png file**

```{r fig.width=7, fig.height=6}
## Make a histogram of the total number of steps taken each day
library(ggplot2)

#setwd(paste0(getwd(),directory))
meanPlot <- ggplot(actSum, aes(x=date, y=totalSteps))
meanPlot <- meanPlot + geom_bar(fill="#00E5EE", stat="identity")
meanPlot <- meanPlot + labs(x = "Date") 
meanPlot <- meanPlot + labs(y = "Total Steps") 
meanPlot <- meanPlot + labs(title="Total number of steps each day")
meanPlot <- meanPlot + theme(plot.title = element_text(lineheight=.8, face="bold"),panel.grid.major = element_blank()
                             ,panel.grid.minor = element_blank())
print(meanPlot)
dev.copy(png, file = "Total Num of steps.png",height = 480, width = 480)
dev.off()
```

*Calculate and report the mean number of steps taken per day*

```{r}
actMean <- mean(actSum$totalSteps)

```

*Calculate and report the median total of steps per day*

```{r}
actMedian <- median(actSum$totalSteps)
```

**What is the average daily activity pattern?**

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r fig.width=7, fig.height=6}
## Create a line plot for the Daily activity pattern by interval

actInterval <- ddply(actData,"interval", summarize, avgSteps=mean(steps, na.rm=TRUE))
head(actInterval)

intPlot <- ggplot(actInterval, aes(x=interval, y=avgSteps))
intPlot <- intPlot + geom_line(colour="red")
intPlot <- intPlot + labs(x = "Interval") 
intPlot <- intPlot + labs(y = "Average Steps") 
intPlot <- intPlot + labs(title="Average number of steps each interval")
intPlot <- intPlot + theme(plot.title = element_text(lineheight=.8, face="bold"),panel.grid.major = element_blank()
                             ,panel.grid.minor = element_blank())
print(intPlot)
dev.copy(png, file = "Avg Num of steps per interval.png",height = 480, width = 480)
dev.off()
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r}
maxSteps <- actInterval[which.max(actInterval$avgSteps),1] #835th interval has the max avg
```

*On average, across all the days in the dataset, the maximum number of steps occurs during interval `r maxSteps`!*


**Imputing missing values**

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
missingVal <- sum(is.na(actData$steps)) #2304

```

*There are `r missingVal` NA values in the dataset.*


Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

*I employed a strategy where I impute the 5-minute interval mean.*

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

actData2 <- actData
actInterval2 <- ddply(actData2,"interval", summarize, avgSteps=mean(steps, na.rm=TRUE))
head(actInterval2)
head(actData2)

for(i in 1:length(actData2$steps)){
    if(is.na(actData2[i,1])){
            print("Yes it matched")
            actData2[i,1] <- actInterval2[which(actInterval2$interval == actData2[i,3]),2]
    }
    else{
        actData2[i,1]<- actData2[i,1]
    }
}
head(actData2)
sum(is.na(actData2$steps))

```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}

actSum2<-ddply(actData2, "date", summarize, totalSteps=sum(steps))
actSum2$date <- as.Date(actSum2$date,format = "%Y-%m-%d")
head(actSum2)
str(actSum2)

meanPlot <- ggplot(actSum2, aes(x=date, y=totalSteps))
meanPlot <- meanPlot + geom_bar(fill="#00E5EE", stat="identity")
meanPlot <- meanPlot + labs(x = "Date") 
meanPlot <- meanPlot + labs(y = "Total Steps") 
meanPlot <- meanPlot + labs(title="Total number of steps each day with Imputed Value")
meanPlot <- meanPlot + theme(plot.title = element_text(lineheight=.8, face="bold"),panel.grid.major = element_blank()
                             ,panel.grid.minor = element_blank())
print(meanPlot)
dev.copy(png, file = "Imputed mean histogram.png",height = 480, width = 480)
dev.off()
actMean2 <- mean(actData2$steps)
actMedian2 <- median(actData2$steps)

```

*The new mean is `r actMean2` compared to the value of `r actMean`. Similarly, the new median is `r actMedian2`, compared to the old value of `r actMedian`.*

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

*This histogram is more centralizied, with more values landing on the average, whereas the tails are mostly not affected.*

**Are there differences in activity patterns between weekdays and weekends?**

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels "weekday"and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}


actData2$dayType <- ifelse(weekdays(as.Date(actData2$date)) %in% c("Saturday","Sunday"), "weekend", "weekday")
actData2$dayType <- as.factor(actData2$dayType)
levels(actData2$dayType)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:
```{r fig.width=7, fig.height=6}
intplot2 <- ddply(actData2,.(interval,dayType), summarize, avgSteps=mean(steps, na.rm=TRUE))
head(intplot2)

intPlot <- ggplot(intplot2, 
            aes(x=interval, 
                y=avgSteps))
intPlot <- intPlot + theme_bw()
intPlot <- intPlot + geom_line(colour="#00BFC4")
intPlot <- intPlot + facet_wrap( ~ dayType, ncol=1)
intPlot <- intPlot + labs(x = "Interval") 
intPlot <- intPlot + labs(y = "Number of steps") 
intPlot <- intPlot + labs(title="Average daily activity")
intPlot <- intPlot + theme(plot.title = element_text(lineheight=.8, face="bold"),
               strip.text.x = element_text(size=12),
               strip.background = element_rect(colour="black",fill="#F6E3CE"))
print(intPlot)
dev.copy(png, file = "Avg Num of steps per interval with Imputed values.png",height = 480, width = 480)
dev.off()

```
